The Two Dualities of Computation: Negative and Fractional Types

### 2012-03-10 rational.tex
intro part 2

\paragraph*{Fractional Types.} 
Consider a similar algebraic manipulation involving fractional types.
\begin{center}
\scalebox{1.1}{
\includegraphics{diagrams/thesis/algebra-wire2.pdf}
}
\end{center}
In the case of negatives, the dotted line indicated the `flow of
control' whereas for fractionals it indicates the `flow of
constraints'. At the heart of logic programming is the idea of
variables that capture constraints. Hence it is useful to trace the
computation corresponding to the algebraic proof above with the
analogy to logic variables in mind.

@@

In the algebra of rational numbers we have that for each {{a}} that is
not equal to {{0}}, {{1 = (1/a) * a}}. In our framework this algebraic
identity corresponds to a type isomorphism {{1 <-> (1/a) * a}}, whose
witness maybe thought of as the `identity function' at {{a}}.  For
example, for the datatype \verb#color = R|G|B#, there are three
possible witnesses for this isomorphism mapping {{()}} to either
{{(1/R,R)}} or {{(1/Gr,Gr)}} or {{(1/B,B)}} with the particular
witness determined by further constraints encountered during the
computation.

 and hence the
witness for the isomorphisms above is generally {{(1/alpha,alpha)}}
where~{{alpha}} is a fresh logic variable.

@@

At this point we have constructed a strange mixing of two functions;
`inputs' of one form `outputs' of the other. If {{(1/alpha1,alpha2)}}
is held by one subcomputation and {{(1/alpha2,alpha1)}} is held by
another subcomputation, these `remixed functions' form a communication
channel between the subcomputations -- unifying {{1/alpha1}} with
\verb|color| {{R}} in one subcomputation, fixes it in the
other. Depending on the choice of the type {{a}}, an arbitrary number
of bits may be communicated. The dual reading of the above, with
additive functions of the form {{0<-> (-a)+b}}, correspond to
control flow and coroutining between two subcomputations -- only one
of which `exists' at any point.

@@

\paragraph*{Contributions and Outline.} 
To summarize, in a computational framework that guarantees that information
is preserved, negative and fractional types provide fascinating mechanisms in
which computations can be sliced and diced, decomposed and recomposed, ran
forwards and backwards, in arbitrary ways. The remainder of the paper
formalizes these informal observations. Specifically our main contributions
are:
\begin{itemize}
\item We extend {{langRev}} our reversible programming language of type
  isomorphisms~\cite{rc2011,infeffects} (see Sec.~\ref{sec:pi}) with a notion
  of negative types, that satisfies the isomorphism {{a + (-a) <-> 0}}. The
  semantics of this extension is expressed by having a ``dual'' evaluator
  that reverses the flow of execution for negative
  values. (Sec.~\ref{sec:neg})
\item We independently extend {{langRev}} with a notion of fractional types,
  that satisfies the isomorphism {{a * (1/a) <-> 1}}. The semantics of this
  extension is expressed by introducing logic variables and a unification
  mechanism to model and resolve the constraints introduced by the fractional
  types. (Sec.~\ref{sec:frac})
\item We combine the above two extensions into one extension that includes
  both negative and fractional types. The resulting language allows any
  rational number to be used as a type. The types satisfy the same familiar
  and intuitive isomorphisms that are satisfied in the mathematical field of
  rational numbers. (Sec.~\ref{sec:rat})
\item We develop programming intuition and argue that negative and fractional
  types ought to be part of the vocabulary of every
  programmer. (Sec.~\ref{sec:prog})
\item We relate our notions of negative and fractional types to previous work
  on continuations. Briefly, we argue that conventional continuations
  conflate negative and fractional components. This observation allows us to
  relate two apparently unrelated lines of work: the first pioneered by
  Filinski~\cite{Filinski:1989:DCI:648332.755574} relating continuations to
  negative types and the second~\cite{Bernardi:2010:CSL:1749618.1749689}
  relating continuations to the fractional types of the Lambek-Grishin
  calculus. (Sec.~\ref{sec:related})
\end{itemize}

>>>>

\paragraph*{Fractional Types.} 
Consider a similar algebraic manipulation involving fractional types.
\begin{center}
\scalebox{1.1}{
\includegraphics{diagrams/thesis/algebra-wire2.pdf}
}
\end{center}
In the case of negatives, the dotted line indicated the `flow of
control' whereas for fractionals it indicates the `flow of
constraints'. At the heart of logic programming is the idea of
variables that capture constraints. Hence it is useful to trace the
computation corresponding to the algebraic proof above, with the
analogy to logic variables in mind.

@@

In the algebra of rational numbers {{1 = (1/a) * a}} if {{a}} is
nonzero.  In our framework this algebraic identity corresponds to a
type isomorphism {{1 <-> (1/a) * a}} (written alternately as
{{a-o*a}}) whose witness maybe thought of as the `identity function'
at the type {{a}}.  For example, for the datatype {{color = R|Gr|B}},
there are three possible witnesses for this isomorphism mapping {{()}}
to either {{(1/R,R)}} or {{(1/Gr,Gr)}} or {{(1/B,B)}} with the
particular witness determined by further constraints encountered
during the computation. More generally, the witness for the
isomorphism above may be represented as {{(1/alpha,alpha)}}
where~{{alpha}} is a fresh logic variable of type {{a}}. With this
view of {{a -o* a}} functions, let us consider the following
manipulations:

@@

At this point we have constructed a strange mixing of two `{{a-o*a}}'
functions; `inputs' of one form `outputs' of the other. If
{{(1/alpha1,alpha2)}} is held by one subcomputation and
{{(1/alpha2,alpha1)}} is held by another subcomputation, these
`remixed functions' form a communication channel between the two
concurrent subcomputations -- unifying {{1/alpha1}} with {{color}}
`{{R}}' in one subcomputation, fixes {{alpha1}} in the other. The type
{{a}} thus takes the role of the type of the communication channel,
indicating how much information can be communicated between the two
subcomputations.  Depending on the choice of the type {{a}}, an
arbitrary number of bits may be communicated.

Dually, the additive reading of the above manipulations correspond to
functions of the form {{a-o+a}}, witnessing isomorphisms of the form
{{0<->(-a)+a}}). The `remixed' additive functions express control flow
transfer between two subcomputations, \emph{only one of which exists}
at any point -- i.e. they capture the essence of coroutines. The
behaviour of additive and multiplicative functions are further
discussed in Sec. \ref{sec:rat}.

@@

\paragraph*{Contributions and Outline.} 
To summarize, in a computational framework that guarantees that information
is preserved, negative and fractional types provide fascinating mechanisms in
which computations can be sliced and diced, decomposed and recomposed, run
forwards and backwards, in arbitrary ways. The remainder of the paper
formalizes these informal observations. Specifically our main contributions
are:
\begin{itemize}
\item We extend {{langRev}} our reversible programming language of type
  isomorphisms~\cite{rc2011,infeffects} (see Sec.~\ref{sec:pi}) with a notion
  of negative types, that satisfies the isomorphism {{a + (-a) <-> 0}}. The
  semantics of this extension is expressed by having a ``dual'' evaluator
  that reverses the flow of execution for negative
  values. (Sec.~\ref{sec:neg})
\item We independently extend {{langRev}} with a notion of fractional types,
  that satisfies the isomorphism {{a * (1/a) <-> 1}}. The semantics of this
  extension is expressed by introducing logic variables and a unification
  mechanism to model and resolve the constraints introduced by the fractional
  types. (Sec.~\ref{sec:frac})
\item We combine the above two extensions into one extension that includes
  both negative and fractional types. The resulting language allows any
  rational number to be used as a type. The types satisfy the same familiar
  and intuitive isomorphisms that are satisfied in the mathematical field of
  rational numbers. (Sec.~\ref{sec:rat})
\item We develop programming intuition and argue that negative and fractional
  types ought to be part of the vocabulary of every
  programmer. (Sec.~\ref{sec:prog})
\item We relate our notions of negative and fractional types to previous work
  on continuations. Briefly, we argue that conventional continuations
  conflate negative and fractional components. This observation allows us to
  relate two apparently unrelated lines of work: the first pioneered by
  Filinski~\cite{Filinski:1989:DCI:648332.755574} relating continuations to
  negative types and the second~\cite{Bernardi:2010:CSL:1749618.1749689}
  relating continuations to the fractional types of the Lambek-Grishin
  calculus. (Sec.~\ref{sec:related})
\end{itemize}
